Mixture Models for Single-Cell Assays with Application to Vaccine Studies

### 2013-05-24 MIMOSA.tex
update paper

\subsection{Single-cell gene expression}
We applied the MIMOSA model to a Fluidigm single-cell gene expression data set. We used the two-sided MIMOSA model because genes could be regulated upward or downward upon stimulation. In order to detect stimulation specific changes of expression, we fit our model to each gene within each stimulation. The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene while preserving inter-subject variability (Figure~\ref{fig:fluidigm} A,B). These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A) and preserved in the posterior estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B). A similar analysis using a two-sided Fisher's exact test and clustering the signed FDR adjust p-values (Figure~\ref{fig:fluidigm} C) does not reveal any stimulation-specific patterns. For an FDR of 10\%, Fisher's exact test identified 13 significant genes, while MIMOSA identified 46 significant genes. The 13 genes identified by Fisher's method were a subset of those identified by MIMOSA.

@@

We examined the performance of the constrained ($p^{(s)}>p^{(u)}$) and unconstrained ($p^{(s)} \ne p^{(u)}$) beta-binomial mixture models via simulations. For the simulation, we used hyper--parameters estimated from a one-sided MIMOSA model fit to ICS data (IL2 univariate) from the primary immunogenicity time point. We simulated data from this constrained model with 200 observations (subjects), a response rate of 60\%, $N=$ 1,000, 5,000, and 10,000 events (cells), with ten independent realizations of data for each $N$. We fit the one-sided MIMOSA model to this data. We evaluated the sensitivity and specificity of the model's ability to correctly identify observations (subjects) from the ``responder'' and ``non-responder'' groups through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change. We repeated this procedure for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A,C). In addition, we examined the nominal \textit{vs.} observed FDR to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} B,D).

For both the constrained and unconstrained simulations, MIMOSA was superior to competing methods, including Fisher's exact test, with respect to sensitivity and specificity even at small values of $N$ (Figure~\ref{fig:simulations} A and C). Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations}, panels B, D).

@@

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the event (cell) proportions drawn from  truncated normal distributions with support $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the hyper--parameters  $\alpha$ and $\beta$ estimated from the ICS data set (see Supplementary Figure 3 panels C and D). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test.

@@

\subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
As a proof-of-concept, we applied our multivariate MIMOSA model for two specific genes in the Fluidigm data, namely CCR7 and GZMK. For this example, $K=2$, and we have four possible combinations.
In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B). Only CCL5 positive cells express BIRC3, and its expression increases upon stimulation. The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations  as in IL2$+$ and/or IFN-$\gamma+$) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions. That is, the number of BIRC3-/CCL5+ cells decreases upon stimulation, while the number of BIRC3+/CCL5+ cells increases. When marginalizing over these cell populations, no difference is apparent in any of the samples. In contrast, the multivariate MIMOSA model tests all polyfunctional cell subpopulations simultaneously, and detects significant differences between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D, black labels). Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.

@@

The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental subjects across multiple conditions (\textit{i.e.}, vaccine responders and non-responders). Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure 3). The univariate MIMOSA model based on the Beta-Binomial distribution allows us to constrain the alternative hypothesis to the case $p^{(s)} > p^{(u)}$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample. This has proven to be useful for the ICS data where stimulation induced changes are expected to be one-sided.

@@

Appendix A, B, and C and Supplementary Figures 1--3 referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:results} and \ref{s:demarkercombos} are available in the Supplementary Material.\vspace*{-8pt}

@@

\begin{figure}
\centering
\includegraphics{TIKZFig2.eps}
%\begin{tikzpicture} [auto,node distance=0cm]
%\node at (0,0) (A) {
%\begin{tikzpicture}
%\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot1.pdf}};
%\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
%\node at (0,1) [font=\small\sffamily] {A} ;
%\end{scope}
%\end{tikzpicture}
%};
%\node [right=of A] (B) {
%\begin{tikzpicture}
%\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot2.pdf}};
%\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
%\node at (-0.1,1.1) [font=\small\sffamily] {B} ;
%\end{scope}
%\end{tikzpicture}
%};
%\node [below=of A] (C) {
%\begin{tikzpicture}
%\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.45\columnwidth]{Figures/FluidigmFisher.pdf}};
%\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
%\node at (-0.05,1) [font=\small\sffamily] {C} ;
%\end{scope}
%\end{tikzpicture}
%};
%\end{tikzpicture}
\caption{Signed posterior probability and difference in empirical  in proportions of single-cells expressing each gene on a 96x96 Fluidigm array. The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control) for a model fit to each stimulation and gene separately. Columns and rows are clustered based on these signed posterior probabilities. B) The empirical differences in proportion of cells expressing a gene in the stimulated vs. control samples. Rows and columns are ordered as in A) for comparison. The traces show the deviations of each cell from zero. Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10). C) Signed posterior probabilities for a model fit per gene but across all stimulations simultaneously. D) Clustering of the signed q-values from Fisher's exact test. Genes selected from Fisher's exact test at the 10\% FDR level. This figure appears in color in the electronic version of this article.}
\label{fig:fluidigm}
\end{figure}

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics{TIKZFig3.eps}
%   \begin{tikzpicture} [auto, node distance=0cm]
%   \node at (0,0) (A) {
%\begin{tikzpicture}
%    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_ROC_5000.pdf}};
%    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
%        \node at (0,1) [font=\small\sffamily] {A} ;
%        \end{scope}
%    \end{tikzpicture}
%    };
%    \node [right=of A] (B) {
%    \begin{tikzpicture}
%    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_FDR_5000.pdf}};
%    \begin{scope}[x={(bar.south east)},y={(bar.north west)}]
%    \node at (0,1) [font=\small\sffamily] {B} ;
%    \end{scope}
%    \end{tikzpicture}
%    };
%        \node [below=of A] (C) {
%    \begin{tikzpicture}
%    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
%    \begin{scope}[x={(c.south east)},y={(c.north west)}]
%    \node at (0,1) [font=\small\sffamily] {C} ;
%    \end{scope}
%    \end{tikzpicture}
%    };
%    \node [right=of C] (D) {
%    \begin{tikzpicture}
%    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
%    \begin{scope}[x={(d.south east)},y={(d.north west)}]
%    \node at (0,1) [font=\small\sffamily] {D} ;
%    \end{scope}
%    \end{tikzpicture}
%    };
%
%\end{tikzpicture}
   \caption{Comparison of positivity detection methods on data simulated from the one-sided and two--sided models. Ten simulations were generated at an $N$ of 50,000 total counts using hyper-parameter estimates from real ICS data (IFN-$\gamma$ expressing CD4+ T-cells stimulated with ENV-1-PTEG from HVTN065) with a five-fold effect size between responder and non-responder components. A) Average ROC curve over the 10 simulated data sets (N=50,000), one--sided B) Average observed and nominal false discovery rate over 10 simulated data sets (N=50,000), one--sided. C) Average ROC curves, two--sided model. D) Average observed and nominal FDR, two--sided model. Curves are shown for MIMOSA, Fisher's exact test, the likelihood ratio test, and log fold-change. Results for MIMOSA fit to a model violating model assumptions, as well as other values of N are in Supplementary Figure 3, and for varying values of I (number of observations) are in Supplementary Figure 4. This figure appears in color in the electronic version of this article.}\label{fig:simulations}
\end{figure}

>>>>

\subsection{Single-cell gene expression}
We applied the MIMOSA model to a Fluidigm single-cell gene expression data set. We used the two-sided MIMOSA model because genes could be regulated upward or downward upon stimulation. In order to detect stimulation specific changes of expression, we fit our model to each gene within each stimulation. The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene while preserving inter-subject variability (Figure~\ref{fig:fluidigm} A). These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A) and closely match the empirical estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B). As a comparison, we also fit the model across all stimulations (Figure~\ref{fig:fluidigm} C). Although this leads to a sensible clustering by stimulation, the observed patterns do not agree as well with the empirical differences. A similar analysis using a two-sided Fisher's exact test and clustering the signed FDR adjust p-values (Figure~\ref{fig:fluidigm} D) does not reveal stimulation-specific patterns, with the exception of CMV pp65 nlv5 stimulation). At an FDR of 10\%, Fisher's exact test identified 13 significant genes, while MIMOSA identified 46 significant genes. The 13 genes identified by Fisher's method were a subset of those identified by MIMOSA.

@@

We examined the performance of the constrained ($p^{(s)}>p^{(u)}$) and unconstrained ($p^{(s)} \ne p^{(u)}$) beta-binomial mixture models via simulations. For the simulation, we used hyper--parameters estimated from a one-sided MIMOSA model fit to ICS data (IL2 univariate) from the primary immunogenicity time point. We simulated data from this constrained model with 200 subjects, a response rate of 50\%, $N=$ 10,000, 25,000, and 50,000 cells, with ten independent realizations of data for each $N$. We fit the one-sided MIMOSA model to this data. We evaluated the sensitivity and specificity of the model's ability to correctly identify  subjects from the ``responder'' and ``non-responder'' groups through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change. We repeated this procedure for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A, C). In addition, we examined the nominal \textit{vs.} observed FDR to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} B,D).

For both the constrained and unconstrained simulations, MIMOSA was superior to competing methods, including Fisher's exact test, with respect to sensitivity and specificity even fewer subjects and smaller cell counts (Supplementary Figures 4 A,B and 5 A,B). Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations}, panels B, D).

@@

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the event (cell) proportions drawn from  truncated normal distributions with support $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the hyper--parameters  $\alpha$ and $\beta$ estimated from the ICS data set (see Supplementary Figure 5 panels C and D). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test.

@@

\subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
As a proof-of-concept, we applied our multivariate MIMOSA model for two specific genes in the Fluidigm data, namely CCR7 and GZMK. For this example, $K=2$, and we have four possible combinations.
In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples, as well as for the sum of the counts in the 3 different positive subsets (Figure~\ref{fig:polyfunctionality} A,B). The number of CCR7-GZML+ cells increases upon stimulation, while the number of CCR7+GZMK- cells decreases. The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all possible polyfunctional cell populations (\textit{i.e.}, cells expressing IL2$+$ or IFN-$\gamma+$), would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions. The multivariate MIMOSA model tests all polyfunctional cell subpopulations simultaneously, and detects significant differences between stimulated and unstimulated conditions in 12 of the 16 subjects (Figure~\ref{fig:polyfunctionality} A, black labels). In contrast, marginalizing over these cell populations, no difference is apparent in any of the samples (Figure~\ref{fig:polyfunctionality} B). Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.

@@

The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental subjects across multiple conditions (\textit{i.e.}, vaccine responders and non-responders). Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure 5). The univariate MIMOSA model based on the Beta-Binomial distribution allows us to constrain the alternative hypothesis to the case $p^{(s)} > p^{(u)}$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample. This has proven to be useful for the ICS data where stimulation induced changes are expected to be one-sided. In some cases, the FDR is not well controlled by any method tested here (Supplementary Figure 1), volcano plots of effect size vs posterior probability of response for different cytokine subsets demonstrate these are subjects at a pre--vaccine time--point with a response at the assay--level (Supplementary Figure 6). Such biological variation is occasionally expected.

@@

Appendix A, B, and C and Supplementary Figures 1--6 referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:results} and \ref{s:demarkercombos} are available in the Supplementary Material.\vspace*{-8pt}

@@

\begin{figure}
\centering
\includegraphics{TIKZFig2.eps}
%\begin{tikzpicture} [auto,node distance=0cm]
%\node at (0,0) (A) {
%\begin{tikzpicture}
%\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot1.pdf}};
%\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
%\node at (0,1) [font=\small\sffamily] {A} ;
%\end{scope}
%\end{tikzpicture}
%};
%\node [right=of A] (B) {
%\begin{tikzpicture}
%\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot2.pdf}};
%\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
%\node at (-0.1,1.1) [font=\small\sffamily] {B} ;
%\end{scope}
%\end{tikzpicture}
%};
%\node [below=of A] (C) {
%\begin{tikzpicture}
%\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.45\columnwidth]{Figures/FluidigmFisher.pdf}};
%\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
%\node at (-0.05,1) [font=\small\sffamily] {C} ;
%\end{scope}
%\end{tikzpicture}
%};
%\end{tikzpicture}
\caption{Signed posterior probability and difference in empirical  in proportions of single-cells expressing each gene on a 96x96 Fluidigm array. The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control) for a model fit to each stimulation and gene separately. Columns and rows are clustered based on these signed posterior probabilities. B) The empirical differences in proportion of cells expressing a gene in the stimulated vs. control samples. Rows and columns are ordered as in A) for comparison. The traces show the deviations of each cell from zero. Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10). C) Signed posterior probabilities for a model fit per gene but across all stimulations simultaneously. D) Clustering of the signed q-values from Fisher's exact test. All genes  were significant at, or below, the 10\% FDR level. This figure appears in color in the electronic version of this article.}
\label{fig:fluidigm}
\end{figure}

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics{TIKZFig3.eps}
%   \begin{tikzpicture} [auto, node distance=0cm]
%   \node at (0,0) (A) {
%\begin{tikzpicture}
%    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_ROC_5000.pdf}};
%    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
%        \node at (0,1) [font=\small\sffamily] {A} ;
%        \end{scope}
%    \end{tikzpicture}
%    };
%    \node [right=of A] (B) {
%    \begin{tikzpicture}
%    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_FDR_5000.pdf}};
%    \begin{scope}[x={(bar.south east)},y={(bar.north west)}]
%    \node at (0,1) [font=\small\sffamily] {B} ;
%    \end{scope}
%    \end{tikzpicture}
%    };
%        \node [below=of A] (C) {
%    \begin{tikzpicture}
%    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
%    \begin{scope}[x={(c.south east)},y={(c.north west)}]
%    \node at (0,1) [font=\small\sffamily] {C} ;
%    \end{scope}
%    \end{tikzpicture}
%    };
%    \node [right=of C] (D) {
%    \begin{tikzpicture}
%    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
%    \begin{scope}[x={(d.south east)},y={(d.north west)}]
%    \node at (0,1) [font=\small\sffamily] {D} ;
%    \end{scope}
%    \end{tikzpicture}
%    };
%
%\end{tikzpicture}
   \caption{Comparison of positivity detection methods on data simulated from the one-sided and two--sided models. Ten simulations were generated at an $N$ of 50,000 total counts using hyper-parameter estimates from real ICS data (IFN-$\gamma$ expressing CD4+ T-cells stimulated with ENV-1-PTEG from HVTN065) with a five-fold effect size between responder and non-responder components. A) Average ROC curve over the 10 simulated data sets (N=50,000), one--sided B) Average observed and nominal false discovery rate over 10 simulated data sets (N=50,000), one--sided. C) Average ROC curves, two--sided model. D) Average observed and nominal FDR, two--sided model. Curves are shown for MIMOSA, Fisher's exact test, the likelihood ratio test, and log fold-change. Results for MIMOSA fit to a model violating model assumptions, as well as other values of N (number of cells) and values of I (number of subjects) are in Supplementary Figures 3 and 4. This figure appears in color in the electronic version of this article.}\label{fig:simulations}
\end{figure}
